Space is the Place: Effects of Continuous Spatial Structure on Analysis of Population Genetic Data

### 2019-11-20 ms/review-responses.tex

\begin{point}{}
    The source of equation \eqref{eqn:p_survival} is not obvious to me. I sort of see how it makes sense, but a little but more intuition or a brief derivation or an illuminating either in the main text or the supplement, would be helpful.
\end{point}

\reply{
    Additional discussion of the model is now provided in the appendix (see "Demographic model") \llname{apx:demographic_model}
}

\begin{point}{}
    The authors use a scaling factor in equation (2) to counteract the increase in fitness of individuals at the edges. Can they provide a figure showing that this is the case. What does ``roughly'' mean on line 164. Perhaps a heatmap of the fitness of individuals across the grid with and without the scaling factor?
\end{point}

\reply{
    Good suggestion -- we have added a supplemental figure (Figure \ref{fig:edge_scaling})
    to show the distribution of individual fitness across the landscape
    with and without our edge-scaling approach.
}

\begin{point}{}
    It would be helpful provide the figure showing that generating mutations during the forward simulations in SLIM is equivalent to applying mutations using msprime on pre-generated trees \revref? It sounds like this procedure would underestimate the variance in the number of mutations, since you remove the effect of random generation time. Is this effect small?
\end{point}

\reply{
    Theory says that any difference here would be extremely small
    and would affect only the variance, not the mean.
    Nonetheless, we have added a figure showing sample site frequency spectra
    generated from a subset of simulations run with SLiM mutations,
    and then using msprime to apply mutations to the same tree sequences with our generation-time scaling approach (Figure \ref{fig:slim_v_msp_sfs}).
    These approaches yield extremely similar spectra.
}

@@

\begin{point}{}
    Can the authors provide a bit more intuition behind the patterns of variation seen in generation time, census population size, and variance in the number of offspring with respect to neighborhood size seen in Figure 2? For example, it is not obvious to me why the census population size, for example, should decline systematically with respect to neighborhood size. Presumably this isn't just due to the local demographic stochasticity. Could the authors briefly interpret the observed patterns or cite appropriate literature?
\end{point}

\reply{
    We agree that this is hard to intuit: there are a lot of factors at work.
    We have added some more discussion of these phenomena \llname{censusscaling}.
}

\begin{point}{}
    \autoref{fig:gwas}D: I am surprised by the extent to which the observed values of $-\log_{10}(p)$ fall below the $y=x$ line. Particularly in the lower right panel for large neighbourhood sizes. I would expect that to be close to panmictic - why are the P-values underdispersed? That seems like a potential bug, or else something weird is going on.
\end{point}

@@

\reply{
    We have checked the code to the best of our abilities and did not find a bug causing the underdispersion. It seems to reflect overcorrection in the regression when using PC coordinates as covariates -- the PCA is capturing some information about the spatial genetic variation which itself covaries only weakly with the phenotype, and as a result we see anomalously low $-log10(p)$ when regressing genotype against phenotype. 
}

\begin{point}{\revref}
    It might be worth citing \citep{haworth2019apparent,},
    who do the proposed test (GWAS for birth location) in UK Biobank to illustrate the population structure.
\end{point}

\reply{
Done - thank you for pointing us to this study. \revref
}

@@

\begin{point}{}
    The analysis and discussion around the effect of GWAS is focused on PCA correction. Do mixed models help at all?
\end{point}

\reply{
    We are also very interested to know how mixed models perform here, but think that adding a second GWAS method would make this section too large for the current paper. We have added a note to \llname{mixedmodelcallout} specifically citing mixed models as alternate methods that may perform better. 
}

\begin{point}{}
    The github link to the code didn't work for me. I assume it will be made public later, but at this point I can't tell whether the code is available/useable.
\end{point}

\reply{
We apologize -- we had forgotten to make the repository public. Now it is.
}

@@

%%%%%%%%%%%%%%%%%%%
\reviewersection{3}

\begin{quote}
    The present study deals with a ``hot topic'' in spatial population genetics. Most
    inferential and descriptive methods in statistical spatial population
    genetic rely on a discrete approximation of space and it is not clear what
    impact this approximation may have when individuals migrate along a
    continuum instead. Spatial patterns in sampling is also another major issue
    which is often simply dismissed, mainly because of the paucity of
    statistical methods to deal with it. This work touches on these important
    issues in a timely manner.
\end{quote}

We are glad you agree this is an important topic to explore!

@@

\begin{quote}
    Although I was enthusiastic about the topic, I was quite disappointed with
    the core of the study, i.e., the forward-in-time simulation of populations in
    continuous space. The field has been struggling with this issue for decades --
    examples of spectacular failures like the Wright-Malecot model (see
    Felsenstein's ``pain in the torus'' article, 1975) or, more recently, the
    ``mugration'' or ``discrete trait analysis'' model in phylodynamics (see De Maio
    et al. 2015) have probably mostly harmed our research field -- that one cannot
    make the economy of using a sound probabilistic model for generating
    geo-referenced genetic data. It does not seem to be the case here
    unfortunately.  
\end{quote}

We were surprised that the reviewer seems to be saying that local, density-mediated control of mortality
doesn't constitute a ``sound probabilisitic model'',
but hope that our additional discussion of the choice,
and citations to the history of this sort of model,
help to better motivate our choice.

@@

\begin{point}{}
    First, the simulation starts with individuals distributed uniformly at random in space. Is there any indication that the three-step algorithm used here maintains this distribution during the course of evolution? If it does not, then is there any stationary regime and how many generations does one need to wait before reaching it? I do appreciate that the competitive interaction term was introduced in order to avoid seeing the ``clumping' of individuals that hampers the Wright-Malecot model. Yet, just because there are no such clusters does not mean that the spatial distribution of individuals reaches a stable regime and that the distribution reached, if any, is reasonable from a biological perspective.
\end{point}

\reply{
    This is a natural question that we asked ourselves while developing the simulation using the built-in visualization tools of the SLiM GUI. We have now added a supplementary figure (Figure \ref{fig:wm_v_dd}) showing the distribution of individuals in our density-dependant spatial model and a continuous-space Wright-Fisher simulation without density dependance. 
    We encourage the reviewer to run our SLiM recipe using the GUI at different parameter values
    to see for themselves how quickly equilibrium is attained --
    we are tempted to include movies of this as supplementary material,
    but that seems excessive.
    Also note that we did \emph{not} include competitive interaction to avoid clumping --
    we included it to make the model biologically realistic
    (so that, for instance, the population does not grow without bound),
    and lack of clumping is a \emph{consequence} of this choice.
    We have clarified this in the text. \revref
}

\begin{point}{}
Second, the demographic process used here involves birth and death of individuals. Does the population survive asymptotically or, like any birth-death process, eventually dies with probability one? In fact, one needs to know a little about the dynamics of the population size to decide whether the corresponding process is reasonable from a biological standpoint.
\end{point}

\reply{
    As in the last point, as part of sanity checking our models
    we monitored the population size over time while with SLiM.
    All simulations maintained asymptotic populations, and none of the runs we started ended because populations crashed.
    Although the population \emph{is} a Markov process with a single accessible absorbing state (death),
    and (since it is not a branching process, so won't grow without bound)
    will eventually die out with probability 1,
    it will take an extremely long time to hit that state.
    We have added a clarifying note about this. \revref
}

@@

\begin{point}{}
    Third, it is not clear what the relationship between the expected lifespan and the probability of survival is. The expected lifespan, L, is first defined as the inverse of the expected number of offspring produced by a parent. The authors also define the probability of survival of a given individual at a given point in space, $p_i$. Hence, the expected lifespan at a point in space (and time) is the mean of a geometric distribution with parameter $p_i$, i.e., $1/p_i$. Now, it is far from being obvious what the relationship between these two approaches for defining the expected lifespan actually is.
\end{point}

\reply{
   We explained the relationship here,
   in a new Appendix (`Demographic Model'; \llname{apx:demographic_model}).
}

\begin{point}{}
    Also, the web page \url{https://github.com/petrelharp/spaceness} does not seem to exist so that I was not able to experiment with the forward-in-time generator used here unfortunately.
\end{point}

\reply{
We apologize -- we had forgotten to make the repository public. Now it is.
}

\begin{point}{}
    All in all, more efforts need to be made here in my opinion to show that the forward-in-time simulations generate sensible outcomes. Sensible in terms of the behavior of the population demography at equilibrium (provided such equilibrium indeed exists) along with that of the spatial distribution of individuals. The authors could provide some guarantee of the good behavior of their model as evidenced from simulations using a broad range of parameter values for generating data. Alternatively, they could elect to use the spatial-Lambda-Fleming-Viot model for their simulations, which, in my opinion would seem the most sensible option given that (1) it is possible to run backward-in-time simulations under this model, thereby saving a lot of computation time and (2) it is a well-studied model with good mathematical and biological properties and (3) it is implemented in a publicly available software program (\url{https://github.com/jeromekelleher/discsim}).
\end{point}

\reply{
    Hopefully, the additional plots 
    and discussion around demographic modeling (notably, the discussion at \llname{demog_disco})
    helps to increase the transparency in this model.
    We agree that demographic realism is a very important consideration,
    and due to the often hard-to-intuit nature of spatial demography,
    simulations need to be carefully sanity checked.
    (Furthermore, what is ``realistic'' for one species will not be for another!)
    We have added an Appendix comparing our model to a stepping-stone \llname{apx:stepping_stone}, 
    because we think these are the most familiar and widely used class of spatial models.
    We find that many features of our model are well approximated by stepping-stone models, and that for statistics like $\theta_{\pi}$ the stepping stone model results approach our continuous space model as the number of demes used to describe the landscape increases. 
}

    We definately think that a comparison to the spatial Lambda-Fleming-Viot
    would be an extremely useful thing to do, for the reasons the reviewer mentions,
    but this would constitute an entirely different study.
    The SLFV model is only known to be an approximation of one particular biologically explicit model
    (of patchy extinction-recolonization)
    and is only conjectured to be a good approximation
    for spatial models more generally.
    To reiterate, the definition of the SLFV was explicitly motivated by a desire to approximate
    spatial models, especially Bolker-Pacala models like ours.
    Therefore, our study is an important prerequisite
    for simulation study of the spatial Lambda-Fleming-Viot,
    as it works through some of the issues in simulating biologically realistic spatial models,
    and describes some of the patterns.

\begin{point}{\autoref{fig:genparams}:}
    I do not understand why the neighborhood size varies to the same extent in the random mating model as it does for the spatial model. For the random mating model, I would have expected the neighborhood size to be equal to the census size since all individuals have the same probability of being a parent of any given offspring. From [the paragraph at \revref], it is clear that the spatial model would converge to the random mating model when the mean parent-offspring distance tends to infinity only if we were to ignore the impact of range edges. I am thus wondering whether the variation of neighborhood size one observes in Fig 2 for the random mating model is just a consequence of border effects. If that is the case, then the authors should state it clearly and try to justify it from a biological perspective.
\end{point}

### 2019-11-20 ms/spaceness.tex
grrr r3

To avoid hard-to-evaluate approximations, we here used forward-time, individual-based simulations.
The question of what regulates real populations has a long history and many answers
\citep[e.g.,][]{lloyd1967crowding,antonovics1980ecological,crawley1990population}, \llabel{ecolcites}
but it is clear that populations must at some point have density-dependent feedback on population size,
or else they would face eventual extinction or explosion.
In the absence of unrealistic global population regulation,
this regulation must be local,
and there are many ways to arrange this \citep{bolker2003spatial}.
In our simulations,
we decrease each individual's probability of survival with increasing local population density,
which shifts reproductive output towards low-density regions 
(and prevents the unrealistic population clustering seen by \citet{Felsenstein1975},
see Supplemental \autoref{fig:slim_has_no_pain}).
Such models have been used extensively in ecological modeling
\citep{durrett1994importance,bolker1997using,law2003population,fournier2004microscopic,champer2019suppression}
but rarely in population genetics, \revpoint{AE}{2}
where to our knowledge implementations of continuous space models
before their availability through SLiM \citep{Haller2019}
have focused on a small number of genetic loci
\citep[e.g.,][]{slatkin1989comparison,barton2002neutral,robledoarnuncio2010isolation,rossine2014espao},
which limits the ability to investigate the impacts of continuous space on genome-wide genetic variation as is now routinely sampled from real organisms. 
By simulating chromosome-scale sequence alignments and complete population histories we are able to treat our simulations as real populations and replicate the sampling designs and analyses commonly conducted on real genomic data.

@@

We ran 400 simulations for the spatial and random-mating models on a square landscape of width $W=50$  \revpoint{AE}{3}
with per-unit carrying capacity $K=5$ (census $N \approx 10,000$), 
average lifetime $L=4$, 
genome size = $10^{8}$, 
recombination rate = $10^{-9}$, 
and drawing $\sigma$ values from a uniform distribution between 0.2 and 4. 
To speed up the simulations and limit memory overhead we used a mutation rate of 0 in SLiM 
and later applied mutations to the tree sequence with msprime's \texttt{mutate} function \citep{Kelleher2016}. 
Because msprime applies mutations proportionally to elapsed time,
we divided the mutation rate of $10^{-8}$ mutations per site per generation
by the average generation time estimated for each value of $\sigma$ (see `Demographic Parameters' below) to convert the rate to units of mutations per site per unit time.
We show that this procedure produces the same site-frequency spectrum as applying mutations directly in SLiM in Figure \ref{fig:slim_v_msp_sfs}. Simulations were run for 1.6 million timesteps (approximately $30N$ generations).

@@

There are a few additional things to note about Figure~\ref{fig:genparams}. 
First, all three quantities are non-monotone with neighborhood size. 
Census size largely declines as neighborhood size increases for both the spatial and random mating models. 
However, for spatial models this decline only begins for neighborhood size $\geq 10$. 
By a neighborhood sizes larger than 100, the spatial and random mating models are indistinguishable from one another, 
a sign that our simulations are performing as expected. 
Census sizes range from $\approx 14,000$ at low $\sigma$ in the random mating model to $\approx 10,000$ for both models when neighborhood sizes approach 1,000. The scaling of census sizes in both random-mating and spatial models appears to be related to two consequences of the spatial competition function: the decline of fitness at range edges, which effectively reduces the habitable area by one $\sigma$ around the edge of the map and so results in a smaller habitable area at high $\sigma$ values; and variation in the equilibrium population density given varying competition radii. For spatial models we see an additional effect in which census size initially rises as neighborhood sizes scale from 2 - 10. This may reflect an Allee effect \citep{Allee1949} in which some individuals are unable to find mates when the mate selection radius is very small.  \llabel{censusscaling}

@@

\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{figures/gwas_summary_loglog.pdf}
    \caption{Impacts of spatially varying environments and isolation by distance on linear regression GWAS. Simulated quantitative phenotypes are determined only by an individual's location and the spatial distribution of environmental factors. In \textbf{A} we show the phenotypes and locations of sampled individuals under four environmental distributions, with transparency scaled to phenotype. As neighborhood size increases a PCA explains less of the total variation in the data (\textbf{B}). Spatially correlated environmental factors cause false positives at a large proportion of SNPs, which is partially but not entirely corrected by adding the first 10 PC coordinates as covariates (\textbf{C}). Quantile-quantile plots in \textbf{D} show inflation of $-\log_{10}(p)$ after PC correction across all simulations and environmental distributions, with colors scaled by the median neighborhood size in each region of q-q space.
}
\label{fig:gwas}
\end{figure*}

@@

We suggest a straightforward alternative for species in which the primary axes of population differentiation is space 
(note this is likely not the case for some modern human populations): 
run a GWAS with spatial coordinates as phenotypes and check for $p$-value inflation or significant associations.
If significant associations with sample locality are observed after correcting for population structure, 
the method is sensitive to false positives induced by spatial structure. This is essentially the approach taken in our ``clinal'' model (though we add normally distributed noise to our phenotypes). This approach has recently been taken with polygenic scores for UK Biobank samples in \cite{haworth2019apparent}, finding that scores are correlated with birth location even in this relatively homogenous sample \llabel{haworthcite}.
Of course, it is possible that genotypes indirectly affect individual locations by adjusting organismal fitness and thus habitat selection across spatially varying environments, but we believe that this hypothesis should be tested against a null of stratification bias inflation rather than accepted as true based on GWAS results.

@@

\section{Supplementary Figures and Tables}
\FloatBarrier

\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{figures/wm_v_dd_maps.pdf}
\caption{Maps of individual locations in a continuous-space Wright-Malécot model with independent dispersal of all individuals (top) and under our continuous space model incorporating density-dependant fitness (bottom). The clustering seen in the top row is the "Pain in the Torus" described by \citet{Felsenstein1975}.}
\label{fig:slim_has_no_pain} 
\end{figure*}
\FloatBarrier

>>>>

To avoid hard-to-evaluate approximations, we here used forward-time, individual-based simulations.
The question of what regulates real populations has a long history and many answers
\citep[e.g.,][]{lloyd1967crowding,antonovics1980ecological,crawley1990population}, \llabel{ecolcites}
but it is clear that populations must at some point have density-dependent feedback on population size,
or else they would face eventual extinction or explosion.
In the absence of unrealistic global population regulation,
this regulation must be local,
and there are many ways to arrange this \citep{bolker2003spatial}.
In our simulations,
we decrease each individual's probability of survival with increasing local population density,
which shifts reproductive output towards low-density regions,
and produces total census sizes that fluctuate around an equilibrium. \revpoint{3}{2}
This also prevents the population clumping seen by \citet{Felsenstein1975}
(Supplemental \autoref{fig:wm_v_dd})). \revpoint{3}{1}
Such models have been used extensively in ecological modeling
\citep{durrett1994importance,bolker1997using,law2003population,fournier2004microscopic,champer2019suppression}
but rarely in population genetics, \revpoint{AE}{2}
where to our knowledge implementations of continuous space models
before their availability through SLiM \citep{Haller2019}
have focused on a small number of genetic loci
\citep[e.g.,][]{slatkin1989comparison,barton2002neutral,robledoarnuncio2010isolation,rossine2014espao},
which limits the ability to investigate the impacts of continuous space on genome-wide genetic variation as is now routinely sampled from real organisms. 
By simulating chromosome-scale sequence alignments and complete population histories we are able to treat our simulations as real populations and replicate the sampling designs and analyses commonly conducted on real genomic data.

@@

We ran 400 simulations for the spatial and random-mating models on a square landscape of width $W=50$  \revpoint{AE}{3}
with per-unit carrying capacity $K=5$ (census $N \approx 10,000$), 
average lifetime $L=4$, 
genome size = $10^{8}$, 
recombination rate = $10^{-9}$, 
and drawing $\sigma$ values from a uniform distribution between 0.2 and 4. 
To speed up the simulations and limit memory overhead we used a mutation rate of 0 in SLiM 
and later applied mutations to the tree sequence with msprime's \texttt{mutate} function \citep{Kelleher2016}. 
Because msprime applies mutations proportionally to elapsed time,
we divided the mutation rate of $10^{-8}$ mutations per site per generation
by the average generation time estimated for each value of $\sigma$ (see `Demographic Parameters' below) to convert the rate to units of mutations per site per unit time. \revpoint{2}{5}
We verify that this procedure produces the same site frequency spectrum as applying mutations directly in SLiM in Figure \ref{fig:slim_v_msp_sfs},
as predicted by theory \citep{ralph2019efficiently}.
Simulations were run for 1.6 million timesteps (approximately $30N$ generations).

@@

There are a few additional things to note about Figure~\ref{fig:genparams}. 
First, all three quantities are non-monotone with neighborhood size. 
Census size largely declines as neighborhood size increases for both the spatial and random mating models. 
However, for spatial models this decline only begins for neighborhood size $\geq 10$. 
Spatial and random mating models are indistinguishable from one another
for neighborhood sizes larger than 100.
Census sizes range from $\approx 14,000$ at low $\sigma$ in the random mating model to $\approx 10,000$ for both models when neighborhood sizes approach 1,000. The scaling of census sizes in both random-mating and spatial models appears to be related to two consequences of the spatial competition function: the decline of fitness at range edges, which effectively reduces the habitable area by one $\sigma$ around the edge of the map and so results in a smaller habitable area at high $\sigma$ values; and variation in the equilibrium population density given varying competition radii.
Furthermore, census size increases in spatial models as neighborhood size increases from 2 to 10.
This may reflect an Allee effect \citep{Allee1949} in which some individuals are unable to find mates when the mate selection radius is very small.  \llabel{censusscaling}

@@

\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{figures/gwas_summary_loglog.pdf}
    \caption{Impacts of spatially varying environments and isolation by distance on linear regression GWAS. Simulated quantitative phenotypes are determined only by an individual's location and the spatial distribution of environmental factors. In \textbf{A} we show the phenotypes and locations of sampled individuals under four environmental distributions, with transparency scaled to phenotype. As neighborhood size increases a PCA explains less of the total variation in the data (\textbf{B}). Spatially correlated environmental factors cause false positives at a large proportion of SNPs, which is partially but not entirely corrected by adding the first 10 PC coordinates as covariates (\textbf{C}). Quantile-quantile plots in (\textbf{D}) show inflation of $-\log_{10}(p)$ after PC correction across all simulations and environmental distributions, with colors scaled by the median neighborhood size in each region of q-q space.
}
\label{fig:gwas}
\end{figure*}

@@

We suggest a straightforward alternative for species in which the primary axes of population differentiation is space 
(note this is likely not the case for some modern human populations): 
run a GWAS with spatial coordinates as phenotypes and check for $p$-value inflation or significant associations.
If significant associations with sample locality are observed after correcting for population structure, 
the method is sensitive to false positives induced by spatial structure. This is essentially the approach taken in our ``clinal'' model (though we add normally distributed noise to our phenotypes). This approach has recently been taken with polygenic scores for UK Biobank samples in \cite{haworth2019apparent}, finding that scores are correlated with birth location even in this relatively homogenous sample. \revpoint{2}{8}
Of course, it is possible that genotypes indirectly affect individual locations by adjusting organismal fitness and thus habitat selection across spatially varying environments, but we believe that this hypothesis should be tested against a null of stratification bias inflation rather than accepted as true based on GWAS results.

@@

\section{Supplementary Figures and Tables}

\begin{figure*}[p]
\centering
\includegraphics[width=\textwidth]{figures/wm_v_dd_maps.pdf}
\caption{Maps of individual locations in a continuous-space Wright-Malécot model with independent dispersal of all individuals (top) and under our continuous space model incorporating density-dependant fitness (bottom). The clustering seen in the top row is the "Pain in the Torus" described by \citet{Felsenstein1975}.}
\label{fig:wm_v_dd} 
\end{figure*}
\FloatBarrier
